Objective-First Design of Nanophotonic Waveguide Couplers

### 2011-10-14 presentations/ob1_wg_paper/paper.tex
Updated with Jelena's edits.

\maketitle
\begin{abstract}
We introduce an ``objective-first'' approach to the design of
    nanophotonic waveguide couplers.
Our method is
    computationally fast (20 minutes), 
    requires no trial-and-error,
    does not require a good starting design,
    can be applied to arbitrary waveguide modes,
    and generates devices with high coupling efficiencies (typically 95\%)
    and small footprints (1-4 square vacuum wavelengths).
\end{abstract}
\ocis{230.7370, 130.3990.}

@@

% Why is waveguide mode conversion important?
\subsection{The importance of waveguide mode conversion}
Optical mode conversion, 
    the efficient transfer of photons from one guided mode to another,
    is a fundamental requirement in nanophotonics.
Efficient conversion between waveguides modes
    is critical in the many cases, including:
    \begin{enumerate}
        \item Coupling to and from optical fiber\cite{fibergrating}, 
        to communicate with the outside world.
    \item Coupling between various nanophotonic waveguides, 
        since different waveguides are best suited for different applications.
        For example, ridge waveguides seem ideal for 
            low-loss transport\cite{ridge},
            but other waveguides, 
            such as photonic crystal waveguides or slot waveguides,
            may be better suited for slow-light\cite{pcslow} 
            or energy-focusing devices\cite{slotfocus}.
    \item Coupling between different materials systems such as
       passive, active\cite{active}, and metallic\cite{metallic} devices. 
    \end{enumerate}
Although efficient waveguide coupling is essential in any nanophotonic system,
    practical methods to design couplers have not yet been developed.

@@

The differences between Eqs.~\ref{eq:adj} and \ref{eq:ob1} are:
\begin{enumerate}
    \item in Eq.~\ref{eq:adj:obj}, we attempt to decrease the design objective,
        while in Eq.~\ref{eq:ob1:con}, the design objective is kept at zero;
    \item in Eq.~\ref{eq:adj:con}, we always satisfy the underlying physics,
        while in Eq.~\ref{eq:ob1:obj}, physics is not satisfied,
            since the physics residual is generally non-zero.
\end{enumerate}

Thus, the fundamental innovation in the objective-first approach
    is simply this:
    we forcibly impose the desired performance on the device at the expense of
    breaking the physics which govern its operation.

@@

The first practical implication of the objective-first approach 
    is that the number of independent variables is increased to include
    both $x$ and $p$.
This occurs because in Eq.~\ref{eq:adj}, 
    the constraint that physics must be satisfied, $g(x,p)=0$, 
    essentially forces $x$ to be a dependentent variable,
    since the choice of $p$ implicitly determines the value of $x$
    (there is generally a one-to-one mapping from $p$ to $x$).
In contrast, Eq.~\ref{eq:ob1} allows both $x$ and $p$ to vary independently,
    because the constraint, $f(x)=0$, is only a function of $x$.

Secondly, the amount of computation needed to enforce the constraint is
    drastically reduced in the objective-first approach.
This is because Eq.~\ref{eq:adj} requires a full solution of $g(x,p)=0$
    (i.e. a full simulation of the structure, $p$) to compute $x$.
In contrast, the constraint $f(x)=0$ in Eq.~\ref{eq:ob1} 
    can often be enforced with a trivial amount of computation (as shown below).

@@

We choose to work in the two-dimensional transverse electric mode,
    which only couples $E_x$, $E_y$, and $H_z$ ($E_z, H_x, H_y = 0$),
    since it is most relevant for on-chip devices.
We choose to use $H_z$ as the field variable, 
    and $\epsilon^{-1}$ (inverse of the permittivity) 
    as the structure variable.
This results in the following representation of the physics residual, 
    based on the time-harmonic electromagnetic wave equation without sources;
    \begin{equation}
    \|g(H_z, \epsilon^{-1})\|^2 = 
    \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu \omega^2 H_z \|^2,
    \end{equation}
    where $\omega$ is the angular frequency,
    and $\mu$ is the permeability of free-space.

For the design objective, we choose a boundary element formulation
    based on $H_z^\text{perfect}$,
    where $H_z^\text{perfect}$ is constructed 
    of the exact input and output waveguide modes at the input and output ports,
    repectively, and of zero-amplitude fields at the unused ports.
The actual form of the design objective is simply, 
    \begin{equation}
    f(H_z) = \begin{bmatrix}
        H_z - H_z^\text{perfect} \\
        \frac{\partial H_z}{\partial n} - 
            \frac{\partial H_z^\text{perfect}}{\partial n}
        \end{bmatrix}_\text{boundary}
        = 0.
    \end{equation}
That is to say,
    the values of $H_z$ and $\partial H_z / \partial n$
    (spatial derivative along normal direction)
    along the device boundary are forced to be 
    those of a device with perfect performance 
    (100\% coupling efficiency).

@@

Finally, as in any method vase on an objective-first approach,
    the physics residual is not guaranteed to decrease to zero.
Thus, it is entirely possible to never achieve 
    a physically realizable field, $H_z$.
In such cases, which are the norm rather than the exception,
    we find that a relatively small residual usually leads to
    fairly good, although imperfect, device performance.

\subsection{Numerical methods used to solve the objective-first design problem}
The design problem is now
    \begin{subequations}\label{eq:act}
    \begin{align} 
    \text{decrease} & \quad  
        \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu \omega^2 H_z \|^2
        \label{eq:act:obj} \\ 
    \text{subject to} & \quad 
        \begin{bmatrix}
        H_z - H_z^\text{perfect} \\
        \frac{\partial H_z}{\partial n} - 
            \frac{\partial H_z^\text{perfect}}{\partial n}
        \end{bmatrix}_\text{boundary}
        = 0.
    \end{align}
    \end{subequations}

This problem contains many local minima (it is non-convex\cite{boydbook});
    however, when either the field ($H_z$) or the structure ($\epsilon^{-1}$)
    variable is considered separately Eq.~\ref{eq:act} has only one minimum
    (it is convex), and can be easily solved using standard methods
    such as employed in our previous work\cite{prevwork}.
We employ such an alternating directions strategy,
    where both $H_z$ and $\epsilon^{-1}$ are solved independently.
This process is extremely inefficient,
    but is employed because the underlying numerical methods
    do not need to be tuned by the user.
We expect considerable improvements in computational efficiency
    when more sophisticated algorithms are applied,
    especially those which can update $H_z$ and $\epsilon^{-1}$ independently.

@@

\section{Results}
The method described above is applied to 
    the design of three different waveguide couplers
    as shown in Figs.~\ref{fig:fiber}, \ref{fig:mode} and \ref{fig:aircore}.
These results demonstrate that our method
    \begin{itemize}
    \item is computationally fast, 
        requiring only 20 minutes on a single-core personal computer;
    \item generates compact devices with footprints of only
        1-4 square vacuum wavelengths;
    \item generates highly efficient devices with  
        typical coupling efficiencies of $95\%$;
    \item can be applied to arbitrary input and output modes,
        as seen from the diverse selection of desired output waveguide modes;
    \item does not require a good initial design.
        The initial design was simply $\epsilon = 9$ everywhere
            (a somewhat arbitrary guess, other values work as well),
            which in the case of Figs.~\ref{fig:mode} and \ref{fig:aircore}
            produces an initially non-functional device 
            (coupling efficiency of $0\%$).

@@

Lastly, note that the couplers in Figs.~\ref{fig:mode} and \ref{fig:aircore}
    must convert between modes of opposing symmetry,
    in which case adiabatic approaches cannot be applied.
Note also that the output waveguide in Fig.~\ref{fig:aircore}
    even operates on a completely different principle 
    (distributed reflection instead of index-guiding).

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/fiber} 
    \caption{Waveguide coupler for a wide, low-index 
        ($\epsilon=2.25$) waveguide. 
        The dielectric structure of the coupler and surrounding input and
        output waveguides is shown on the left, while the simulation
        validating our results is shown on the right.
        The coupler converts $96.3\%$ of the input power to the
        designated output mode.
        The device is extremely compact, 
        convering only $36 \times 66$ grid points,
        where the vacuum wavelength is 42 grid points.
        Computation time was 20 minutes on a personal computer.}
    \label{fig:fiber}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/mode-conv} 
    \caption{Coupler that converts the fundamental waveguide mode to the
        second-order waveguide mode
        (a movie of the design progress is included as a
        supplementary material).
        This problem is quite difficult since the two modes are of 
        opposite symmetry.
        For example, adiabatic approaches cannot be applied to this case.
        However, our method produces a device 
        (which has the same dimensions and vacuum wavelength as Fig. 1) 
        with a coupling efficiency of $95.5\%$. 
        Computation time was 20 minutes on a personal computer.
        }
    \label{fig:mode}
    \end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{quickfig/air-core}
    \caption{Coupler to an air-core mode.
        Here, not only are the modes of opposite symmetry,
        but the output waveguide operates on a fundamentally different
        principle (distributed reflection) than the input waveguide 
        (index guided).
        The device still achieves an efficiency of $83.3\%$, demonstrating the
        versatility of our method.
        The vacuum wavelength is 25 grid points, 
        while the device footprint is still $36 \times 66$ grid points.
        Computation time was 20 minutes on a personal computer.
        }
        \label{fig:aircore}
\end{figure}

>>>>

\maketitle
\begin{abstract}
We introduce an ``objective-first'' approach to the design of
    nanophotonic waveguide couplers.
Our method is
    computationally fast (20 minutes on a single-core personal computer), 
    requires no trial-and-error,
    does not require guessing a good starting design,
    can be applied to arbitrary waveguide modes,
    and generates devices with high coupling efficiencies (typically 95\%)
    and small footprints (1-4 square vacuum wavelengths).
\end{abstract}
\ocis{230.7370, 130.3990.}

@@

% Why is waveguide mode conversion important?
\subsection{The importance of waveguide mode conversion}
Optical mode conversion, 
    the efficient transfer of photons from one guided mode to another,
    is a fundamental requirement in nanophotonics.
Efficient conversion between waveguides modes
    is critical in the many cases, including:
    \begin{enumerate}
        \item Coupling to and from optical fiber\cite{fibergrating}, 
        to communicate with the outside world.
    \item Coupling between various nanophotonic waveguides, 
        since different waveguides are best suited for different applications.
        For example, ridge waveguides seem ideal for 
            low-loss transport\cite{ridge},
            but other waveguides, 
            such as photonic crystal waveguides or slot waveguides,
            may be better suited for slow-light\cite{pcslow} 
            or nonlinear optical devices based on 
            localized field intensities\cite{slotfocus}.
    \item Coupling between different materials systems such as
       passive, active\cite{active}, and metallic\cite{metallic} devices. 
    \end{enumerate}
Although efficient waveguide coupling is essential in any nanophotonic system,
    practical methods to design couplers have not yet been developed.

@@

The differences between Eqs.~\ref{eq:adj} and \ref{eq:ob1} are:
\begin{enumerate}
    \item in Eq.~\ref{eq:adj:obj}, we attempt to decrease the design objective,
        while in Eq.~\ref{eq:ob1:con}, the design objective is kept at zero
        (i.e. we force the optimization to satisfy the design objective);
    \item in Eq.~\ref{eq:adj:con}, we always satisfy the underlying physics,
        while in Eq.~\ref{eq:ob1:obj}, physics is not necessarily satisfied,
            since the physics residual is generally non-zero
            (although the optimization is directed at minimizing this residual).
\end{enumerate}

Thus, the fundamental innovation in the objective-first approach
    is simply this:
    we forcibly impose the desired performance on the device at the expense of
    not perfectly satisfying the physical equation which governs its operation.

@@

The first practical implication of the objective-first approach 
    is that the number of independent variables is increased to include
    both $x$ and $p$.
This occurs because in Eq.~\ref{eq:adj}, 
    the constraint that physics must be satisfied, $g(x,p)=0$, 
    essentially forces $x$ to be a dependent variable,
    since the choice of $p$ implicitly determines the value of $x$
    (there is generally a one-to-one mapping from $p$ to $x$).
In contrast, Eq.~\ref{eq:ob1} allows both $x$ and $p$ to vary independently,
    because the constraint, $f(x)=0$, is only a function of $x$.

Secondly, the amount of computation needed to enforce the constraint is
    drastically reduced in our objective-first approach.
This is because Eq.~\ref{eq:adj} requires a full solution of $g(x,p)=0$
    (i.e. a full simulation of the structure, $p$) to compute $x$.
In contrast, the constraint $f(x)=0$ in Eq.~\ref{eq:ob1} 
    can often be enforced with a trivial amount of computation (as shown below).

@@

We choose to work in the two-dimensional transverse electric mode,
    which only couples $E_x$, $E_y$, and $H_z$ ($E_z, H_x, H_y = 0$),
    since it is most relevant for on-chip devices.
We choose to use $H_z$ as the field variable ($x$ in Eq.~\ref{eq:ob1}), 
    and $\epsilon^{-1}$ (inverse of the permittivity) 
    as the structure variable $p$.
This results in the following representation of the physics residual, 
    based on the time-harmonic electromagnetic wave equation without sources;
    \begin{equation}
    \|g(H_z, \epsilon^{-1})\|^2 = 
    \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu_0 \omega^2 H_z \|^2,
    \end{equation}
    where $\omega$ is the angular frequency,
    and $\mu_0$ is the permeability of free-space.

For the design objective, we choose a boundary-value formulation
    based on $H_z^\text{perfect}$,
    where $H_z^\text{perfect}$ is constructed 
    of the exact input and output waveguide modes at the input and output ports
    of the optimization region (where the coupler will be placed),
    repectively, and of zero-amplitude fields at the unused ports,
    as illustrated in Fig.~\ref{fig:intro}.

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{quickfig/intro} 
    \caption{Boundary-value formulation of the design objective.
        The values of $H_z^\text{perfect}$, 
            defined along the dashed box surrounding the design area
            (coupling structure), 
            are shown in orange (color online).
        The values of $H_z^\text{perfect}$ along the top and bottom edges
            of the dashed box are set to zero.
        In this schematic, the fundamental and second-order waveguide modes
            have been chosen as the input and output modes respectively.}
    \label{fig:intro}
\end{figure}

The mathematical form of the design objective is simply, 
    \begin{equation}
    f(H_z) = \begin{bmatrix}
        H_z - H_z^\text{perfect} \\
        \frac{\partial H_z}{\partial n} - 
            \frac{\partial H_z^\text{perfect}}{\partial n}
        \end{bmatrix}_\text{boundary}
        = 0.
    \end{equation}
That is to say,
    the values of $H_z$ and $\partial H_z / \partial n$
    (spatial derivative along normal direction)
    along the device boundary are forced to be 
    those of a device with perfect performance 
    (100\% coupling efficiency).

@@

Finally, as in any method based on an objective-first approach,
    the physics residual is not guaranteed to decrease to zero.
Thus, it is entirely possible to never achieve 
    a physically realizable field, $H_z$.
In such cases, which are the norm rather than the exception,
    we find that a relatively small residual usually leads to
    fairly good, although imperfect, device performance.

\subsection{Numerical methods used to solve the objective-first design problem}
The design problem is now
    \begin{subequations}\label{eq:act}
    \begin{align} 
    \text{decrease} & \quad  
        \| \nabla \times \epsilon^{-1} \nabla \times H_z - 
            \mu_0 \omega^2 H_z \|^2 \label{eq:act:obj} \\ 
    \text{subject to} & \quad 
        \begin{bmatrix}
        H_z - H_z^\text{perfect} \\
        \frac{\partial H_z}{\partial n} - 
            \frac{\partial H_z^\text{perfect}}{\partial n}
        \end{bmatrix}_\text{boundary}
        = 0.
    \end{align}
    \end{subequations}

This problem contains many local minima (it is non-convex\cite{boydbook});
    however, when either the field ($H_z$) or the structure ($\epsilon^{-1}$)
    variable is considered separately, Eq.~\ref{eq:act} has only one minimum
    (it is convex), and can be easily solved using standard methods
    such as employed in our previous work\cite{prevwork}.
We employ such an alternating directions strategy,
    where both $H_z$ and $\epsilon^{-1}$ are solved independently.
This process is extremely inefficient,
    but is employed because the underlying numerical methods
    do not need to be tuned by the user.
We expect considerable improvements in computational efficiency
    when more sophisticated algorithms are applied,
    especially those which can update $H_z$ and $\epsilon^{-1}$ independently.

@@

\section{Results}
We apply our method to three practically interesting problems 
    that are hard to solve using other available methods: 
    \begin{enumerate}
    \item a coupler between waveguides of different refractive index and width,
    \item a coupler between waveguide modes of different order and symmetry, and
    \item a coupler between waveguides that confine light 
        using different principles 
        (index guided vs. distributed Bragg reflection guided), 
        i.e., between a slab waveguide and a photonic crystal fiber. 
    \end{enumerate}

The corresponding designs as well as the boundary values which make up the
    respective design objectives ($H_z^\text{perfect}$) 
    are shown in Figs.~\ref{fig:fiber}, \ref{fig:mode} and \ref{fig:aircore}.
As initial designs, we uniformly choose $\epsilon = 9$ throughout 
    the whole computational region.
These results demonstrate that our method
    \begin{itemize}
    \item is computationally fast, 
        requiring only 20 minutes on a single-core personal computer;
    \item generates compact devices with footprints of only
        1-4 square vacuum wavelengths;
    \item generates highly efficient devices with  
        typical coupling efficiencies of $95\%$;
    \item can be applied to arbitrary input and output modes,
        as seen from the diverse selection of desired output waveguide modes;
    \item does not require a good initial design.
        The initial design was simply $\epsilon = 9$ everywhere
            (a somewhat arbitrary guess, other values work as well),
            which in the case of Figs.~\ref{fig:mode} and \ref{fig:aircore}
            produces an initially non-functional device 
            (coupling efficiency of $0\%$).

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{quickfig/fiber} 
    \caption{Coupler from a narrow, high-index ($\epsilon=12.25$) waveguide to
        a wide, low-index ($\epsilon=2.25$) waveguide. 
        On the left, the dielectric structure of the coupler 
        and surrounding input and output waveguides are shown.
        The $H_z^\text{perfect}$ boundary values used as 
        the design objective are plotted on the left as well
        (top and bottom boundary values were set to zero).
        On the right, we show the simulation results used
        to compute the performance of the device.
        The coupler converts $96.3\%$ of the input power to the
        designated output mode.
        The device is also extremely compact, 
        convering only $36 \times 66$ grid points,
        where the vacuum wavelength is 42 grid points.
        Computation time was 20 minutes on a personal computer.}
    \label{fig:fiber}
\end{figure}
\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{quickfig/mode-conv} 
    \caption{Coupler that converts the fundamental waveguide mode to the
        second-order waveguide mode
        (a movie of the design progress is included as a
        supplementary material).
        This problem is quite difficult since the two modes are of 
        opposite symmetry.
        For example, adiabatic approaches cannot be applied to this case.
        However, our method produces a device 
        (which has the same dimensions and vacuum wavelength as Fig. 1) 
        with a coupling efficiency of $95.5\%$. 
        Computation time was 20 minutes on a personal computer.
        }
    \label{fig:mode}
    \end{figure}
\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{quickfig/air-core}
    \caption{Coupler between a dielectric slab waveguide to 
        an air-core waveguide.
        Here, not only are the modes of opposite symmetry,
        but the output waveguide operates on a fundamentally different
        principle (distributed reflection) than the input waveguide 
        (index guided).
        The device still achieves an efficiency of $83.3\%$, demonstrating the
        versatility of our method.
        The vacuum wavelength is 25 grid points, 
        while the device footprint is still $36 \times 66$ grid points.
        Computation time was 20 minutes on a personal computer.
        }
        \label{fig:aircore}
\end{figure}

\section{Conclusion}
We develop a fundamentally new approach to designing physical structures,
    which we term ``objective-first'', 
    in that we choose to satisfy the design objective 
    even above satisfying the physical equation which governs its operation.
We show that such an approach drastically reduces the
    amount of computation required per iteration, and 
    performs well even with a non-functional initial design.

We then apply an objective-first approach to the design of 
    three practical nanophotonic waveguide couplers which are difficult, 
    at best, to solve with existing methods. 
We show that our method produces
    high-efficiency designs ($\sim 95\%$ efficiency) 
    in small footprints ($\sim 1$ square vacuum wavelength),
    is computationally fast (20 minutes on a single-core personal computer), and
    does not require trial-and-error, or even 
    a good starting design.\\
